---
title: "02.5.2.eems_overall"
output: html_document
---
# libraries
```{r}
library(rnaturalearth)
library(ggplot2)
library(sf)
library(terra)
library(dplyr)
library()
```

# installs
installing feems into a conda environment called feems_py310
```{bash}
conda create -n feems_py310 python=3.10 -y
conda activate feems_py310

# Use conda-forge for compiled wheels consistently
mamba install -c conda-forge numpy=1.23.* scipy pandas scikit-learn matplotlib statsmodels \
  shapely pyproj fiona geopandas -y

# Then install feems + pandas-plink
pip install feems pandas-plink
install -c bioconda feems -c conda-forge
```

# input files
## PLINK files
These have already been made for the population structure stuff, so I'm using those:
/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.all.recode

## coords file
Need the sample order from the fam file
```{bash}
rsync -avzP ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/BANS.all.recode.fam /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/02.5.delineate_ESUs
```

```{r}
fam <- read.table("BANS.all.recode.fam",
  stringsAsFactors = FALSE
)

colnames(fam)[1:2] <- c("FID", "IID")

meta <- read.csv("../../data/BANS_all_sample_data_shifted.csv") %>% select(BGP_ID, Long, Lat) %>% rename(IID=BGP_ID)

coords_df <- left_join(fam, meta, by="IID")

#project coordinates to better system
coords_sf <- st_as_sf(
  coords_df,
  coords = c("Long", "Lat"),
  crs = 4326
)

coords_proj <- st_transform(coords_sf, 5070)

#get coords input file
xy <- st_coordinates(coords_proj)

write.table(
  xy,
  file = "BANS.coords.txt",
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)
```

```{bash}
rsync -avzP /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/02.5.delineate_ESUs/BANS.coords.txt ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/
```
## boundary shape file
```{r}
#north america without political boundaries
map <- ne_states(country = c("United States of America", "Canada"), 
                 returnclass = 'sf')
na_union <- st_union(map)

#BANS breeding range
breeding.sf <- st_read("../../data/spatial_files/banswa_range_2023/banswa_range_2023.gpkg") %>% filter(season=="breeding") %>% st_transform(crs(map))

#clip BANS range to just NA and make a single, clean polygon
breeding_na <- st_intersection(breeding.sf, na_union)
breeding_na <- st_make_valid(breeding_na)

breeding_outer <- st_union(breeding_na)

#project to match the coords file
breeding_outer_proj <- st_transform(breeding_outer, 5070)

#get outer boundary
outer_coords <- st_coordinates(breeding_outer_proj)[, 1:2]

write.table(
  outer_coords,
  "BANS_breeding_outer.txt",
  row.names = FALSE,
  col.names = FALSE
)

#make grids within the polygon (I believe this is what gets filled in with migratory surface values)
grid <- st_make_grid(
  breeding_outer_proj,
  cellsize = 75000,   # 75 km
  square = TRUE
)

grid <- st_sf(geometry = grid)
grid <- st_intersection(grid, breeding_outer_proj)

st_write(grid, "BANS_breeding_grid_75km.shp", delete_dsn = TRUE)


plot(st_geometry(na_union), col = "grey90")
plot(st_geometry(breeding_na), col = "steelblue", add=T)
```

Problem: Feems wants a continuous surface, not the disjointed mess that is this birds breeding range. So, I need to simplify/blend this range...
```{r}
#north america without political boundaries
map <- ne_states(country = c("United States of America", "Canada"), 
                 returnclass = 'sf')
na_union <- st_union(map)

#BANS breeding range
breeding.sf <- st_read("../../data/spatial_files/banswa_range_2023/banswa_range_2023.gpkg") %>% filter(season=="breeding") %>% st_transform(crs(map))

breeding_outer <- st_union(breeding.sf)

#project to match the coords file
breeding_outer_proj <- st_transform(breeding_outer, 5070)

# outer_poly <- st_union(st_buffer(breeding_outer_proj, 100000))
# outer_poly <- st_make_valid(outer_poly)

#############
#remove texas
st_bbox(breeding_outer_proj)

clip_box <- st_as_sfc(st_bbox(c(
  xmin = -4514069.6,
  ymin = 321028.6 + 600000,  # move north ~300 km
  xmax = 2798098.4,
  ymax = 6133850.1
), crs = st_crs(breeding_outer_proj)))

rng_clipped <- st_intersection(breeding_outer_proj, clip_box)
rng_clipped <- st_union(rng_clipped)

#hull <- st_convex_hull(breeding_outer_proj)
hull <- st_concave_hull(rng_clipped, allow_holes = F, ratio = 0.25)


#############

ggplot() +
  geom_sf(data = breeding_outer_proj, fill = "red", color="red")+
  geom_sf(data=hull, fill="yellow", alpha=0.5)
```

```{r}
#get outer boundary
outer_coords <- st_coordinates(hull)[, 1:2]

write.table(
  outer_coords,
  "BANS_breeding_outer.txt",
  row.names = FALSE,
  col.names = FALSE
)

#make grids within the polygon (I believe this is what gets filled in with migratory surface values)
cellsize_m <- 75000

grid_sq <- st_make_grid(hull, cellsize = cellsize_m, square = TRUE) %>%
  st_as_sf() %>%
  st_set_crs(st_crs(hull))

# Keep only squares whose centroid is inside the hull polygon
grid_cent <- st_centroid(grid_sq)
inside <- st_within(grid_cent, hull, sparse = FALSE)[,1]
grid_sq <- grid_sq[inside, , drop = FALSE]

# sanity check: ensure all are POLYGON (not MULTIPOLYGON)
table(st_geometry_type(grid_sq))

plot(st_geometry(grid_sq))

st_write(grid_sq, "BANS_breeding_grid_75km.shp", delete_dsn = TRUE)
```

```{bash}
rsync -avzP /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/02.5.delineate_ESUs/BANS_breeding_* ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/
```
# OLD---another method for input boundaries
```{r}
coords_sf <- st_as_sf(coords, coords = c("Long", "Lat"), crs = 5070) %>%
  st_make_valid()
# concave hull around coordinates
hull <- st_concave_hull(coords_sf, allow_holes = F, ratio = 0.25)
hull <- st_make_valid(hull)

breeding.sf <- st_read("../../data/spatial_files/banswa_range_2023/banswa_range_2023.gpkg") %>% filter(season=="breeding") %>% st_transform(crs(map))

# project and clean geometries
range_proj <- st_transform(breeding.sf, 5070) %>% st_buffer(0)
hull_proj  <- st_transform(hull, 5070) %>% st_buffer(0)

#remove texas
st_bbox(range_proj)

clip_box <- st_as_sfc(st_bbox(c(
  xmin = -4514069.6,
  ymin = 321028.6 + 1500000,  # move north ~700 km
  xmax = 2798098.4,
  ymax = 6133850.1
), crs = st_crs(range_proj)))

rng_clipped <- st_intersection(range_proj, clip_box)
range.hull <- st_concave_hull(rng_clipped, allow_holes = F, ratio = 0.25)

# combine points poplygon and range map
combined_proj <- st_union(range.hull, hull_proj) %>%
  st_make_valid() %>%
  st_union()

# #buffer around new polygon
# combined <- st_buffer(combined_proj, 50000) %>%  # 50 km
#   st_make_valid()
# 
# parts <- st_collection_extract(combined, "POLYGON")
# if (length(parts) > 1) {
#   parts <- st_cast(parts, "POLYGON", warn = FALSE)
# }
# combined_poly <- parts[which.max(st_area(parts))]
# 
# combined_poly <- st_simplify(combined_poly, dTolerance = 30000, preserveTopology = TRUE) %>%
#   st_make_valid()
# 
# # give crs
combined_final <- st_transform(combined_proj, 5070)

map <- ne_states(
  country = c("United States of America", "Canada"),
  returnclass = "sf")

ggplot() +
  geom_sf(data = map, fill = "white", color = "black", size = 0.2) +
  geom_sf(data = combined_final, fill = "grey70", color = "black", alpha = 0.5) +
  geom_sf(data = coords_sf, color = "blue", size = 2)

#get outer boundary
outer_coords <- st_coordinates(combined_final)[, 1:2]

write.table(
  outer_coords,
  "BANS_breeding_outer2.txt",
  row.names = FALSE,
  col.names = FALSE
)

#make grids within the polygon (I believe this is what gets filled in with migratory surface values)
grid <- st_make_grid(
  combined_final,
  cellsize = 75000,   # 75 km
  square = TRUE
)

grid <- st_sf(geometry = grid)
grid <- st_intersection(grid, combined_final)

plot(st_geometry(grid))

st_write(grid, "BANS_breeding_grid_75km2.shp", delete_dsn = TRUE)
```

```{bash}
rsync -avzP /Users/ericarobertson/Desktop/BANS_adaptive_units/analysis/02.5.delineate_ESUs/BANS_breeding_*2* ericacnr@colostate.edu@login.rc.colorado.edu:/scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc/
```

## another method
```{r}
library(sf)
library(dplyr)
library(rnaturalearth)

# ----------------------------
# Inputs
# ----------------------------
coords_df <- coords  # must have Long/Lat columns (lon/lat)
range_ll <- st_read("../../data/spatial_files/banswa_range_2023/banswa_range_2023.gpkg") %>%
  filter(season == "breeding") %>%
  st_make_valid()

# If coords are lon/lat, this MUST be 4326 first:
coords_sf_ll <- st_as_sf(coords_df, coords = c("Long", "Lat"), crs = 4326) %>%
  st_make_valid()

# Project everything to an equal-area CRS in meters (you chose 5070)
coords_sf <- st_transform(coords_sf_ll, 5070)
range_proj <- st_transform(range_ll, 5070) %>% st_make_valid() %>% st_union()

# ----------------------------
# (Optional) Clip out Texas-ish by moving ymin north.
# Tune yshift_m until TX is excluded but everything else remains.
# ----------------------------
bb <- st_bbox(range_proj)
yshift_m <- 250000  # 250 km north; adjust 150k–400k as needed

clip_bb <- st_bbox(c(
  xmin = as.numeric(bb["xmin"]),
  ymin = as.numeric(bb["ymin"]) + yshift_m,
  xmax = as.numeric(bb["xmax"]),
  ymax = as.numeric(bb["ymax"])
), crs = st_crs(range_proj))

clip_box <- st_as_sfc(clip_bb)
range_proj <- st_intersection(range_proj, clip_box) %>% st_make_valid() %>% st_union()

# ----------------------------
# Concave hull around samples (projected)
# ----------------------------
hull_proj <- st_concave_hull(st_union(coords_sf), ratio = 0.25, allow_holes = FALSE) %>%
  st_make_valid()

# ----------------------------
# Combine range + hull, then buffer slightly (50 km)
# ----------------------------
combined <- st_union(range_proj, hull_proj) %>%
  st_make_valid() %>%
  st_union()

combined <- st_buffer(combined, 50000) %>%  # 50 km
  st_make_valid()

# ----------------------------
# Force to a SINGLE polygon (largest piece)
# ----------------------------
parts <- st_collection_extract(combined, "POLYGON")
if (length(parts) > 1) {
  parts <- st_cast(parts, "POLYGON", warn = FALSE)
}
combined_poly <- parts[which.max(st_area(parts))]

# ----------------------------
# Simplify to prevent millions of vertices (critical!)
# Tune dTolerance: start 10–25 km.
# ----------------------------
combined_poly <- st_simplify(combined_poly, dTolerance = 20000, preserveTopology = TRUE) %>%
  st_make_valid()

# Optional: segmentize after simplify (keeps shape but caps segment length)
# combined_poly <- st_segmentize(combined_poly, dfMaxLength = 25000)

# Quick sanity checks
print(st_geometry_type(combined_poly))
cat("Outer vertex count:", nrow(st_coordinates(combined_poly)), "\n")
```
# feems python script adjusted
So this is an adjusted version of the one that is found here https://github.com/NovembreLab/feems?tab=readme-ov-file. ChatGPT tidied it up and made it compatible with my inputs above.

```{python}
print("STEP 2: Prepare graph inputs")
    print("=" * 70)

    outer, edges, grid, _ = prepare_graph_inputs(
        coord=coord,
        ggrid=grid_path,
        translated=args.translated,
        buffer=args.buffer,
        outer=outer_input
    )

    print(f"[INFO] Grid nodes: {len(grid)}")
    print(f"[INFO] Graph edges: {len(edges)}")

    print("\n" + "=" * 70)
    print("STEP 3: Initialize SpatialGraph")
    print("=" * 70)

    sp_graph = SpatialGraph(genotypes, coord, grid, edges, scale_snps=True)
    print("[INFO] SpatialGraph initialized.")

    # CV (optional)
    best_lamb = args.lamb
    best_lamb_q = args.lamb_q

    if args.run_cv:
        print("\n" + "=" * 70)
        print("STEP 4: Cross-validation (joint)")
        print("=" * 70)

        LAMB_GRID = parse_logspace_triplet(args.lamb_grid)
        LAMB_Q_GRID = parse_logspace_triplet(args.lamb_q_grid)

        print(f"[INFO] CV grid: {len(LAMB_GRID)} lamb x {len(LAMB_Q_GRID)} lamb_q")
        print(f"[INFO] n_folds={args.n_folds}")

        cv_err = run_cv_joint(
            sp_graph,
            LAMB_GRID,
    
```

# feems as a slurm job
```{bash}
#!/bin/bash
#
#SBATCH --job-name=feems
#SBATCH --output=feems.%j.out
#SBATCH --error=feems.%j.err
#SBATCH -t 23:00:00
#SBATCH --partition=amilan
#SBATCH --qos=normal
#SBATCH --nodes=1
#SBATCH --ntasks-per-node 11
#SBATCH --mem=40G
#SBATCH --mail-type=BEGIN,END,FAIL
#SBATCH  --mail-user=ericacnr@colostate.edu


set -x
source ~/.bashrc

conda activate feems_py310

python h_run_feems.py \
  --workdir /scratch/alpine/ericacnr@colostate.edu/BANS/02.5.neutral_pop_struc \
  --plink_prefix BANS.all.recode \
  --coords BANS.coords.txt \
  --outer BANS_breeding_outer.txt \
  --grid BANS_breeding_grid_75km.shp \
  --outdir feems_run1 \
  --lamb 2.0 \
  --lamb_q 10.0 \
  --outlier_frac 0.01 \
  --nedges 3 \
  --top 5
```
