---
title: "RDA"
author: "Erica Robertson"
date: "2025-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="~/Desktop/BANS/BANS_adaptive_units/")
```

```{r, label="packages"}
library(pegas)
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(RColorBrewer)
library(ggpubr)
library(vegan)
library(robust)
library(ggVennDiagram)
library(cowplot)
library(corrplot)
library(data.table)
library(tidyverse)
library(qvalue)
library(adegenet)
```


I want to run the rda on individual genotypes for all 1.4ish million snps. So I need to provide the R rda() command with the genotypes and the enviornmental predictors for each individual. 
```
plink formatting
mamba activate plink pass-maf-0.05-SNP-5miss-294ind-rm-scaf-lth10-gtgl-impute4.1-snpid
plink --bfile --keep losh.267ind.33pop.iidfid.txt --recode A --aec --out pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid


## this is actually not the cut I used, I used an awk script that does the same.
cut -d$'\t' -f7- pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid.raw > pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid-noheader.raw

```
load allele freq, env, pop data
```{r}
pop <- fread("~/Downloads/pop.order.txt", header=F) %>% rename(pop = V1)
all.freq <- fread("data/losh.33pop.frq.strat.50k_A.txt") ## instead of running on allele frequency per population, try running on individual genotypes...

## you have two options, remove the header in plink before you import (recommended)
genotypes <- fread("data/pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid-noheader.raw")

## adegenet::read.PLINK will do it for you, but requires converting to a df...
gen <- read.PLINK("data/pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid.raw")
gen.imp <- as.data.frame(gen)


indorder <- fread("data/pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid.fam") %>% rename(BGP_ID = V1) %>% dplyr::select(BGP_ID)

#env <- fread("data/losh.clim.hii.poporder.txt") avg per pop
#env <- fread("../env-variables/LOSH.264ind.33pop.worldclim.txt") %>% select(BGP_ID, matches("bio_(18|8|7|2)")) %>% #rename_with(~ c("bio18", "bio8", "bio7", "bio2"), matches("bio_(18|8|7|2)")) 

# env <- fread("../env-variables/LOSH.264ind.33pop.worldclim.txt") %>%
#   select(BGP_ID, 
#          bio18 = wc2.1_30s_bio_18,
#          bio8  = wc2.1_30s_bio_8,
#          bio7  = wc2.1_30s_bio_7,
#          bio2  = wc2.1_30s_bio_2,
#          bio15 = wc2.1_30s_bio_15,
#          bio3  = wc2.1_30s_bio_3) 


env <- read_delim("../env-variables/LOSH.264ind.33pop.worldclim.txt", delim = "\t") %>% select(BGP_ID, wc2.1_30s_bio_18, wc2.1_30s_bio_8, wc2.1_30s_bio_7, wc2.1_30s_bio_2, wc2.1_30s_bio_15, wc2.1_30s_bio_3) %>% rename(bio18 = wc2.1_30s_bio_18, bio8 = wc2.1_30s_bio_8, bio7 = wc2.1_30s_bio_7, bio2 = wc2.1_30s_bio_2, bio15 = wc2.1_30s_bio_15, bio3 = wc2.1_30s_bio_3)

env <- left_join(indorder, env, by = "BGP_ID")

env$BGP_ID == indorder$BGP_ID

env %>% select(-"BGP_ID") %>% write.table("data/LOSH.264.ind.33pop.climtop6.txt", row.names = F, quote = F, sep = "\t")

env %>% select(-"BGP_ID", bio18, bio8, bio7, bio2) %>% write.table("data/LOSH.264.ind.33pop.climtop4.txt", row.names = F, quote = F, sep = "\t")


#

## read this in in the rda.R script
env2 <- fread("data/LOSH.264.ind.33pop.climtop6.txt", sep = "\t")



```

```{r}
# source("./src/rdadapt.R")
# rdadapt_env_unconstrained <- rdadapt(RDA_env_unconstrained, 2)
# thres_env <- 0.01/length(rdadapt_env_unconstrained$p.values)
# 
# 
# ## Identifying the outliers for the simple RDA
# outliers_unconstrained <- data.frame(Loci = colnames(all.freq)[which(rdadapt_env_unconstrained$p.values<thres_env)], p.value = rdadapt_env_unconstrained$p.values[which(rdadapt_env_unconstrained$p.values<thres_env)], contig = unlist(lapply(strsplit(colnames(all.freq)[which(rdadapt_env_unconstrained$p.values<thres_env)], split = "_"), function(x) x[1])))
# outliers_unconstrained <- outliers_unconstrained[order(outliers_unconstrained$contig, outliers_unconstrained$p.value),]
# outliers_rdadapt_env_unconstrained <- as.character(outliers_unconstrained$Loci[!duplicated(outliers_unconstrained$contig)])
```

###plink pca
```
plink --bfile pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid --pca 10 header --out pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid --aec 
```
```{r}
# load required libraries
library(adespatial) # for spatial eigenfunction analysis like dbMEM
library(sf) # for handling spatial features (coordinates, projections, etc.)
library(vegan) # for multivariate analysis (e.g., RDA, PCA, variable selection)
library(data.table)

indorder <- fread("data/pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid.fam") %>% rename(BGP_ID = V1) %>% dplyr::select(BGP_ID)

# read in plink pca
plink_pca <- read.table("/Users/holdenfox/Downloads/pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid.eigenvec", header = TRUE)

# clean and rename: drop 'FID' and rename 'IID' to 'BGP_ID' to match other metadata
plink_pca <- plink_pca %>% dplyr::select(-FID) %>% rename(BGP_ID = IID)

## mems
# read in coordinates from metadata file and filter for individuals in indorder
coords <- read_csv("../../meta/losh_master_092924.csv") %>%
  filter(BGP_ID %in% indorder$BGP_ID) %>% 
  dplyr::select(BGP_ID, Long, Lat)

# reorder coordinates to match PCA data
coords <- coords[(match(indorder$BGP_ID, coords$BGP_ID)),]

coords_sf <- st_as_sf(coords, coords = c("Long", "Lat"), crs = 4326)

gd <- geosphere::distm(as(coords_sf, "Spatial"), fun = geosphere::distHaversine)
dist <- as.dist(gd)

# generate moran's eigenvector maps (MEMs) based on coordinates
mem <- adespatial::dbmem(dist, MEM.autocor = "all")

#moran_test <- moran.randtest(mem)

# rda to assess how much MEMs explain genetic variation (PC1–PC3)
mem_rda <- vegan::rda(plink_pca[,c("PC1", "PC2", "PC3")] ~ ., as.data.frame(mem[,1:10]))
summary(mem_rda)

# create a null (intercept-only) rda model as a baseline
mem_rda_full <- vegan::rda(plink_pca[,c("PC1", "PC2", "PC3")] ~ 1, as.data.frame(mem[,1:10]))

# stepwise forward selection of mem variables based on adjusted R-square and significance
selmem <- vegan::ordiR2step(mem_rda_full, mem_rda, Pin = .01)
summary(selmem)
# extract the names of the selected mem variables
selmem <- names(selmem$terminfo$ordered)

# subset the mem matrix to include only the mems that explain population structure
mem_popstr <- mem[,selmem]

#saveRDS(mem_popstr, file = "data/losh.mems.popstr.rds")

```


```{r}
coords_mems_pc <- cbind(coords, mem, plink_pca[,c("PC1", "PC2", "PC3")])

map <- rnaturalearth::ne_states(country = c("United States of America", "Canada", "Mexico"), returnclass = 'sf')
# plot the populations you defined. use plotly to zoom around easy
mem4_map<- ggplot(map) + geom_sf() + geom_point(data = coords_mems_pc, aes(x = Long, y = Lat, color = MEM4, BGP_ID = BGP_ID)) +
coord_sf(xlim = c(-125, -70), ylim = c(23, 58)) + scale_color_viridis_c()
ggsave("mem4_map.pdf", mem4_map, width = 8, height = 6)


plot(coords_mems_pc[,c("PC1", "PC2", "PC3", paste0("MEM",1:9))])
```

One thing the paper that I saw in the paper recommending this did afterwards was select MEMs from the final set (selmem) which are uncorrelated with your selected environmental variables as well. You can do that like:
```{r}
corselmem <- cor(mem[,selmem], pred) # remove any correlated
corselmem <- rownames(corselmem)[which(rowSums(abs(corselmem) >= 0.7) > 0)]
uncorselmem <- mem[,selmem]
uncorselmem <- as.data.frame(uncorselmem)[,-which(colnames(uncorselmem) %in% corselmem), drop = FALSE]
#saveRDS(uncorselmem, file = "data/losh.mems.popstr.uncor.rds")
```

I got annoyed with trying to plot and so stats on all data so i'm going to do a subset of 10,000 snps like brenna does.
```{r}
library(data.table)
library(adegenet)
library(adegenet)

# Load PLINK raw file as a genlight object
# plink_data <- read.PLINK("data/pass-maf-0.05-SNP-5miss-267ind-33pop-rm-scaf-lth10-gtgl-impute4.1-snpid.raw")
# # convert genlight object to a genotype matrix
# geno_matrix <- as.matrix(plink_data)  # This extracts SNP genotypes
# # get SNP names
# snp_names <- colnames(geno_matrix)
# # ensure there are enough SNPs to sample
# num_snps_to_sample <- min(10000, length(snp_names)) ##there is
# # randomly sample 10,000 SNPs (or fewer if there aren't that many)
# set.seed(123)
# selected_snps <- sample(snp_names, num_snps_to_sample)
# # subset the genotype matrix
# geno_subset <- geno_matrix[, selected_snps]
# # convert back to a genlight object
# plink_subset <- new("genlight", geno_subset)
# # save the subsetted SNP data (as a CSV or other format)
# write.table(as.data.frame(geno_subset), "subset_10000_snps.csv", sep = ",", quote = FALSE, row.names = TRUE)

subset <- read.csv("~/Downloads/subset_10000_snps.csv")
env <- fread("data/LOSH.264.ind.33pop.climtop6.txt")
pop <- fread("../losh.33pop.txt", header = F) %>% dplyr::select(-V2) %>% rename(BGP_ID = V1, ClimGroup = V3)
bgp_ids <- data.frame(BGP_ID = rownames(subset))
pop <- pop %>% slice(match(bgp_ids$BGP_ID, pop$BGP_ID)) ## make ind same order as snpfile & env data
env <- cbind(pop, env)

##modify data types like brenna did...

# Convert BGP_ID to character (to avoid factor-related issues)
env[, BGP_ID := as.character(BGP_ID)]

# Convert ClimGroup to a factor if it's categorical
env[, ClimGroup := as.factor(ClimGroup)]

# Convert bio2 to numeric (to match bio18, bio8, bio7)
env[, bio2 := as.numeric(bio2)]

#reordering?
# env[, BGP_ID := as.character(BGP_ID)]
# rownames(subset) <- as.character(rownames(subset))
# 
# env <- env[match(rownames(subset), env$BGP_ID), ]

all(rownames(subset) == env$BGP_ID)

pred <- subset(env, select=c(bio18, bio8, bio7, bio2, bio15, bio3))

pred_mems <- cbind(pred, as.data.frame(mem_popstr))

pred_pc <- cbind(pred, plink_pca %>% dplyr::select(PC1, PC2, PC3))

losh.subset.ind.rda <- rda(subset ~ ., data=pred, scale=T)

# constrain using mems that explain population structure
losh.subset.ind.rda.memps <- rda(subset ~ bio18 + bio8 + bio7 + bio2 + Condition(MEM1 + MEM2 + MEM4), data = pred_mems, scale = T)
losh.subset.ind.rda.memps.uncor <- rda(subset ~ bio18 + bio8 + bio7 + bio2 + Condition(MEM2 + MEM4 + MEM3), data = pred_mems, scale = T)
losh.subset.ind.rda.mem1 <- rda(subset ~ bio18 + bio8 + bio7 + bio2 + Condition(MEM1), data = pred_mems, scale = T)


losh.subset.ind.rda.pc1 <- rda(subset ~ bio18 + bio8 + bio7 + bio2 + Condition(PC1), data = pred_pc, scale = T)
losh.subset.ind.rda.pc1.2 <- rda(subset ~ bio18 + bio8 + bio7 + bio2 + Condition(PC1 + PC2), data = pred_pc, scale = T)

RsquareAdj(losh.subset.ind.rda)
RsquareAdj(losh.subset.ind.rda.memps)
RsquareAdj(losh.subset.ind.rda.memps.uncor)
RsquareAdj(losh.subset.ind.rda.mem1)
RsquareAdj(losh.subset.ind.rda.pc1)

summary(eigenvals(losh.subset.ind.rda, model = "constrained"))
summary(eigenvals(losh.subset.ind.rda.memps, model = "constrained"))

signif.full <- anova.cca(losh.subset.ind.rda.memps, parallel=getOption("mc.cores")) # default is permutation=999
signif.full

vif.cca(losh.subset.ind.rda.memps)




plot(losh.subset.ind.rda, scaling=3)          # default is axes 1 and 2
plot(losh.subset.ind.rda, choices = c(1, 3), scaling=3)

plot(losh.subset.ind.rda.memps, scaling=3)
plot(losh.subset.ind.rda.memps, choices = c(1, 3), scaling=3) 

plot(losh.subset.ind.rda.mem1, scaling=3)
plot(losh.subset.ind.rda.mem1, choices = c(1, 3), scaling=3) 

plot(losh.subset.ind.rda.pc1, scaling = 3)
plot(losh.subset.ind.rda.pc1, choices = c(1, 3), scaling=3) 


plot(losh.subset.ind.rda, scaling=3) 
plot(losh.subset.ind.rda.memps, scaling=3)
plot(losh.subset.ind.rda.memps.uncor)
plot(losh.subset.ind.rda.mem1, scaling=3)
plot(losh.subset.ind.rda.pc1, scaling = 3)
```

```{r}
library(RColorBrewer)
clim_groups <- unique(env$ClimGroup)
eco <- factor(env$ClimGroup, levels = clim_groups)  # Ensure factor levels match
bg <- colorRampPalette(brewer.pal(min(12, length(clim_groups)), "Paired"))(length(clim_groups))


region_order <- c(
  "WA_1", "OR_1", "CA_2", "CA_3", "NV_2", "MT_2", "CO_1", "UT_1", "AZ_1",
  "TX_1", "TX_2", "OK_1", "AR_1", "MS_1", "AL_1", "KY_2", "TN_1", "TN_2", "IN_1", "IL_1",
  "Ontario_1", "Ontario_2",
  "NC_1", "NC_2", "VA_1", "VA_2", "WV_1", "FL_1", "LA_1",
  "Manitoba_1", "Saskatchewan_1", "Alberta_1", "SD_2"
)

eco <- factor(env$ClimGroup, levels = region_order)

# generate colors based on the reordered factor
bg <- colorRampPalette(brewer.pal(min(12, length(unique(eco))), "Paired"))(length(unique(eco)))

# check colors
barplot(rep(1, 33), col = bg, border = NA)


# factor.levels <- c("WA_1", "OR_1", "CA_2", "CA_3", "NV_2", "CO_1", "UT_1", "AZ_1",
#                    "TX_1", "TX_2", "OK_1", "AR_1", "MS_1", "AL_1", 
#                    "KY_2", "TN_1", "TN_2", "IN_1", "IL_1",
#                    "Ontario_1", "Ontario_2",
#                    "Manitoba_1", "Saskatchewan_1", "Alberta_1", "SD_2", "MT_2",
#                    "NC_1", "NC_2", "VA_1", "VA_2", "WV_1", "FL_1", "LA_1"
#                    )
# 
# # Define the correct number of colors for each region
# W.cols <- brewer.pal(n = 9, name = "Reds")[2:9]          
# M.cols <- brewer.pal(n = 6, name = "Greys")[2:7]  
# Midwest.cols <- brewer.pal(n = 9, name = "Blues")[c(1, 3, 5, 7, 9)]  
# Ont.cols <- rep("turquoise4", 2)
# Plains.cols <- brewer.pal(n = 5, name = "Greens")[c(3:7)]  
# SE.cols <- brewer.pal(n = 7, name = "Purples")[1:7] 
# 
# bg <- c(W.cols, M.cols, Midwest.cols, Ont.cols, Plains.cols, SE.cols)
# names(bg) <- factor.levels
# 
# barplot(rep(1, 33), col = bg, border = NA)

# W.cols <- setNames(brewer.pal(n = 9, name = "Reds")[2:9], 
#                    c("WA_1", "OR_1", "CA_2", "CA_3", "NV_2", "CO_1", "UT_1", "AZ_1"))
# M.cols <- setNames(brewer.pal(n = 9, name = "Greys")[c(1, 3, 5, 7, 9)], 
#                    c("TX_1", "TX_2", "OK_1", "AR_1", "MS_1"))
# Midwest.cols <- setNames(brewer.pal(n = 9, name = "Blues")[c(1, 3, 5, 7, 9)],
#                          c("KY_2", "TN_1", "TN_2", "IN_1", "IL_1"))
# Ont.cols <- setNames(rep("turquoise4", 2), c("Ontario_1", "Ontario_2"))
# 
# Plains.cols <- setNames(brewer.pal(n = 7, name = "Greens")[c(3:7)],
#                         c("Manitoba_1", "Saskatchewan_1", "Alberta_1", "SD_2", "MT_2"))
# SE.cols <- setNames(brewer.pal(n = 9, name = "Purples")[2:9],
#                     c("NC_1", "NC_2", "VA_1", "VA_2", "WV_1", "FL_1", "LA_1", "AL_1"))

W.cols <- setNames(brewer.pal(n = 9, name = "Reds")[2:9], 
                   c("WA_1", "OR_1", "CA_2", "CA_3", "NV_2", "CO_1", "UT_1", "AZ_1"))

# Modify M.cols: Remove AL_1
M.cols <- setNames(brewer.pal(n = 6, name = "Greys")[2:6],  # one fewer color
                   c("TX_1", "TX_2", "OK_1", "AR_1", "MS_1"))

# Midwest
Midwest.cols <- setNames(brewer.pal(n = 9, name = "Blues")[c(1, 3, 5, 7, 9)],
                         c("KY_2", "TN_1", "TN_2", "IN_1", "IL_1"))

# Ontario
Ont.cols <- setNames(rep("turquoise4", 2), c("Ontario_1", "Ontario_2"))

# Plains
Plains.cols <- setNames(brewer.pal(n = 7, name = "Greens")[3:7],
                        c("Manitoba_1", "Saskatchewan_1", "Alberta_1", "SD_2", "MT_2"))

# SE: Add AL_1 at the end
SE.cols <- setNames(brewer.pal(n = 9, name = "Purples")[2:9],
                    c("NC_1", "NC_2", "VA_1", "VA_2", "WV_1", "FL_1", "LA_1", "AL_1"))

# Combine all named color vectors
bg <- c(W.cols, M.cols, Midwest.cols, Ont.cols, Plains.cols, SE.cols)

# Reorder bg to match factor.levels explicitly
bg <- bg[factor.levels]

# Plot to check
barplot(rep(1, length(bg)), col = bg, border = NA)

```

```{r}
#pdf("losh.subset.ind.rda.12.pdf")
par(mar=c(5, 5, 4, 8))
plot(losh.subset.ind.rda.memps, type="n", scaling=3)  # Set up empty plot
#points(losh.subset.ind.rda.memps, display="species", pch=20, cex=0.7, col="gray32", scaling=3) # SNPs
points(losh.subset.ind.rda.memps, display="sites", pch=21, cex=1.3, col="gray32", scaling=3, bg=bg[eco]) # Individuals
text(losh.subset.ind.rda.memps, scaling=3, display="bp", col="#0868ac", cex=1)  # Predictors
#legend("bottomright", legend=levels(eco), bty="n", col="gray32", pch=21, cex=1, pt.bg=bg)
#legend("bottomright", legend=levels(eco), bty="n", col="gray32", pch=21, cex=0.6, pt.bg=bg)
#dev.off()

# plot(losh.subset.ind.rda.memps, type="n",choices=c(1,3), scaling=3)  # Set up empty plot
# points(losh.subset.ind.rda.memps, display="species", pch=20, cex=0.7, col="gray32", scaling=3) # SNPs
# points(losh.subset.ind.rda.memps, display="sites", pch=21, cex=1.3, col="gray32", scaling=3, bg=bg[eco]) # Individuals
# text(losh.subset.ind.rda.memps, scaling=3, display="bp", col="#0868ac", cex=1)  # Predictors
# #legend("bottomright", legend=levels(eco), bty="n", col="gray32", pch=21, cex=1, pt.bg=bg)
# legend("bottomright", legend=levels(eco), bty="n", col="gray32", pch=21, cex=0.6, pt.bg=bg)


# mapping them too
coords_env <- left_join(coords, env, by = "BGP_ID")

# plotly::ggplotly(ggplot(map) + geom_sf() + geom_point(data = coords_env, aes(x = Long, y = Lat, color = ClimGroup, BGP_ID = BGP_ID)) + scale_fill_manual(values = bg, name = "ClimGroup") +
# coord_sf(xlim = c(-140, -65), ylim = c(23, 60)))

p <- ggplot(map) +
  geom_sf(fill = "gray95", color = "white") +
  geom_sf(
    data = coords_env, aes(fill = ClimGroup),
    shape = 21, size = 3, color = "black", stroke = 0.2
  ) +
  scale_fill_manual(values = bg, name = "ClimGroup") +
  coord_sf(xlim = c(-140, -65), ylim = c(23, 60)) +
  theme_minimal()
ggsave("losh.subset.ind.rda.memps.map.pdf", p, width = 8, height = 6)

# Interactive plot with tooltips
plotly::ggplotly(p, tooltip = c("text", "fill"))
p

par(mar=c(5, 5, 4, 8))
plot(losh.subset.ind.rda.memps, type="n", scaling=3)  # Set up empty plot
#points(losh.subset.ind.rda.memps, display="species", pch=20, cex=0.7, col="gray32", scaling=3) # SNPs
points(losh.subset.ind.rda.memps, display="sites", pch=21, cex=1.3, col="gray32", scaling=3, bg=bg[eco]) # Individuals
text(losh.subset.ind.rda.memps, scaling=3, display="bp", col="#0868ac", cex=1)

```








# following the correct brenna tutorial
```{r}
losh.ind.rda.4 <- readRDS("data/losh.ind.rda.top4clim.memps.rds")
losh.ind.rda.6 <- readRDS("data/losh.ind.rda.top6clim.memps.rds")
RsquareAdj(losh.ind.rda.6)
summary(eigenvals(losh.ind.rda.6, model = "constrained"))
screeplot(losh.ind.rda.6)
```

```{r}
RsquareAdj(losh.ind.rda)
```
```
$r.squared
[1] 0.06129724

$adj.r.squared
[1] 0.0469659
```
The adjusted r-square explains ~ 5% of the variance

```{r}
summary(eigenvals(losh.ind.rda.3, model = "constrained"))
```
```
Importance of components:
                           RDA1      RDA2      RDA3      RDA4
Eigenvalue            5.523e+04 1.558e+04 7.949e+03 5.580e+03
Proportion Explained  6.548e-01 1.848e-01 9.424e-02 6.615e-02
Cumulative Proportion 6.548e-01 8.396e-01 9.338e-01 1.000e+00
```
The proportion of variance explained by each PC is: RDA1: 65.48%, RDA2: 18.48%, RDA3: 9.42%, RDA4: 6.62%


```{r}

## Not sure how to get these lines to run. Seem to stall out??

# signif.full <- anova.cca(losh.ind.rda, parallel=getOption("mc.cores")) # default is permutation=999
# signif.full

# signif.axis <- anova.cca(losh.ind.rda, by="axis", parallel=getOption("mc.cores"))
# signif.axis

plot(losh.ind.rda.4, scaling=3)
```


```{r}
# pop.xy <- read_csv("../../meta/losh_master_092924.csv") %>%
#   filter(BGP_ID %in% pop$BGP_ID) %>% left_join(pop, by = "BGP_ID") %>% dplyr::select(BGP_ID, ClimGroup, Lat, Long)
# 
# group_centroids <- pop.xy %>%
#   group_by(ClimGroup) %>%
#   summarize(
#     mean_lat = mean(Lat, na.rm = TRUE),
#     mean_long = mean(Long, na.rm = TRUE)
#   )
# 
# geo_pca <- prcomp(group_centroids[, c("mean_long", "mean_lat")], scale. = TRUE)
# 
# # use the first principal component (PC1) as the gradient axis
# group_centroids$geo_score <- geo_pca$x[, 1]
# 
# # arrange by PC1 (diagonal axis)
# group_centroids <- group_centroids %>%
#   arrange(geo_score)
# 
# # create a color palette with as many unique ClimGroups as needed
# geo_colors <- colorRampPalette(brewer.pal(min(12, nrow(group_centroids)), "YlGnBu"))(nrow(group_centroids))
# 
# # assign colors to ClimGroups
# group_centroids$color <- geo_colors
# 
# # create a named vector for easy mapping later
# clim_colors <- setNames(group_centroids$color, group_centroids$ClimGroup)
# 
# pop.xy$color <- clim_colors[pop.xy$ClimGroup]
# 
# barplot(rep(1, length(clim_colors)), col = clim_colors, border = NA,
#         names.arg = names(clim_colors), las = 2, cex.names = 0.7)
# 
# geo_order <- group_centroids$ClimGroup
# eco <- factor(env$ClimGroup, levels = geo_order)
# 
# # create bg color vector matching individuals in env to geographic colors
# bg <- clim_colors[as.character(eco)]


pop.xy <- read_csv("../../meta/losh_master_092924.csv") %>%
  filter(BGP_ID %in% pop$BGP_ID) %>%
  left_join(pop, by = "BGP_ID") %>%
  select(BGP_ID, ClimGroup, Lat, Long)

# Calculate centroids per ClimGroup
group_centroids <- pop.xy %>%
  group_by(ClimGroup) %>%
  summarize(
    mean_lat = mean(Lat, na.rm = TRUE),
    mean_long = mean(Long, na.rm = TRUE),
    .groups = "drop"
  )

# Arrange by **mean latitude** for north-south gradient
group_centroids <- group_centroids %>%
  arrange(desc(mean_lat))  # Descending so that northern groups are at the top

# Generate geographic color gradient
geo_colors <- colorRampPalette(brewer.pal(min(12, nrow(group_centroids)), "YlGnBu"))(nrow(group_centroids))

# Assign colors to ClimGroups
group_centroids$color <- geo_colors
clim_colors <- setNames(group_centroids$color, group_centroids$ClimGroup)

# Map colors to each population point
pop.xy$color <- clim_colors[pop.xy$ClimGroup]

# Optional: show color bar
barplot(rep(1, length(clim_colors)), col = clim_colors, border = NA,
        names.arg = names(clim_colors), las = 2, cex.names = 0.7)

# Create ordered factor for plotting
geo_order <- group_centroids$ClimGroup
eco <- factor(env$ClimGroup, levels = geo_order)

# Assign colors to individuals for plotting
bg <- clim_colors[as.character(eco)]
```

# make this plot colorful
```{r}
all(rownames(env) == rownames(losh.ind.rda.6)) ##order matches if TRUE
## it does

plot(losh.ind.rda.6, type = "n", scaling = 3)  # Empty RDA plot
#points(losh.ind.rda.6, display = "species", pch = 20, cex = 0.7, col = "gray32", scaling = 3)  # SNPs
points(losh.ind.rda.6, display = "sites", pch = 21, cex = 1.3, col = "gray32", scaling = 3, bg = bg[eco])  # Individuals
text(losh.ind.rda.6, scaling = 3, display = "bp", col = "#0868ac", cex = 1)  # Predictors
#legend("bottomright", legend = levels(eco), bty = "n", col = "gray32", pch = 21, pt.bg = bg, cex = 0.6)

```

## plot pop region colors
```{r}
W.cols <- setNames(brewer.pal(n = 9, name = "Reds")[2:9], 
                   c("WA_1", "OR_1", "CA_2", "CA_3", "NV_2", "CO_1", "UT_1", "AZ_1"))
# Modify M.cols: Remove AL_1
M.cols <- setNames(brewer.pal(n = 6, name = "Greys")[2:6],  # one fewer color
                   c("TX_1", "TX_2", "OK_1", "AR_1", "MS_1"))
# Midwest
Midwest.cols <- setNames(brewer.pal(n = 9, name = "Blues")[c(1, 3, 5, 7, 9)],
                         c("KY_2", "TN_1", "TN_2", "IN_1", "IL_1"))
# Ontario
Ont.cols <- setNames(rep("turquoise4", 2), c("Ontario_1", "Ontario_2"))
# Plains
Plains.cols <- setNames(brewer.pal(n = 7, name = "Greens")[3:7],
                        c("Manitoba_1", "Saskatchewan_1", "Alberta_1", "SD_2", "MT_2"))
# SE: Add AL_1 at the end
SE.cols <- setNames(brewer.pal(n = 9, name = "Purples")[2:9],
                    c("NC_1", "NC_2", "VA_1", "VA_2", "WV_1", "FL_1", "LA_1", "AL_1"))
# Combine all named color vectors
bg <- c(W.cols, M.cols, Midwest.cols, Ont.cols, Plains.cols, SE.cols)
# Reorder bg to match factor.levels explicitly
#bg <- bg[factor.levels]
# Plot to check
barplot(rep(1, length(bg)), col = bg, border = NA)

p <- ggplot(map) +
  geom_sf(fill = "gray95", color = "white") +
  geom_point(
    data = coords_env,
    aes(x = Long, y = Lat, fill = ClimGroup, text = BGP_ID),
    shape = 21, size = 3, color = "black", stroke = 0.2
  ) +
  scale_fill_manual(values = bg, name = "ClimGroup") +
  coord_sf(xlim = c(-140, -65), ylim = c(23, 60)) +
  theme_minimal()


#eco <- meta$EcoRegion[match(rownames(losh.ind.rda.6$sites), meta$SampleID)]

# Interactive plot with tooltips
plotly::ggplotly(p, tooltip = c("text", "fill"))
p

pdf("losh.ind.rda.6.memps.pdf")
par(mar=c(5, 5, 4, 8))
plot(losh.ind.rda.6, type="n", scaling=3)  # Set up empty plot
#points(losh.subset.ind.rda.memps, display="species", pch=20, cex=0.7, col="gray32", scaling=3) # SNPs
points(losh.ind.rda.6, display="sites", pch=21, cex=1.3, col="gray32", scaling=3, bg=bg[eco]) # Individuals
text(losh.ind.rda.6, scaling=3, display="bp", col="#0868ac", cex=1)
dev.off()
```





# select outlier snps
```{r}
load.rda <- scores(losh.ind.rda.6, choices=c(1:3), display="species")

hist(load.rda[,1], main="Loadings on RDA1")
hist(load.rda[,2], main="Loadings on RDA2")
hist(load.rda[,3], main="Loadings on RDA3")

outliers <- function(x,z){
  lims <- mean(x) + c(-1, 1) * z * sd(x)   #find loadings +/-z sd from mean loading     
  x[x < lims[1] | x > lims[2]]             #locus names in these tails
}

cand1 <- outliers(load.rda[,1],2) # 75998
cand2 <- outliers(load.rda[,2],2) # 63347
cand3 <- outliers(load.rda[,3],2) # 63725

##Find total # candidates
ncand <- length(cand1) + length(cand2) + length(cand3)

df.cand1 <- cbind.data.frame(rep(1,times=length(cand1)), names(cand1), unname(cand1))
df.cand2 <- cbind.data.frame(rep(2,times=length(cand2)), names(cand2), unname(cand2))
df.cand3 <- cbind.data.frame(rep(3,times=length(cand3)), names(cand3), unname(cand3))

colnames(df.cand1) <- colnames(df.cand2) <- colnames(df.cand3) <- c("axis","snp","loading")

df.cand <- rbind(df.cand1, df.cand2, df.cand3)
df.cand$snp <- as.character(df.cand$snp)

df.cand$snp <- gsub("-", ".", df.cand$snp)

pred <- subset(env, select=c(bio18, bio8, bio7, bio2, bio15, bio3))

##Add environmental correlations to candidate snps
foo <- matrix(nrow=(ncand), ncol=6)  #6 columns for 6 predictors
colnames(foo) <- c("bio18", "bio8", "bio7","bio2", "bio15", "bio3")
pred2 <- pred[,1:6]

for (i in 1:length(df.cand$snp)) {
  nam <- df.cand[i,2]
  snp.gen <- subset[[nam]]
  foo[i,] <- apply(pred2,2,function(x) cor(x,snp.gen))
}

cand <- cbind.data.frame(df.cand,foo)

length(cand$snp[duplicated(cand$snp)])
foo <- cbind(cand$axis, duplicated(cand$snp)) 
table(foo[foo[,1]==1,2]) # no duplicates on axis 1
table(foo[foo[,1]==2,2]) # 18202 duplicates on axis 2
table(foo[foo[,1]==3,2]) # 18787 duplicates on axis 3
cand <- cand[!duplicated(cand$snp),] # remove duplicate detections

cols <- as.numeric(ncol(cand))
for (i in 1:length(cand$snp)) {
  bar <- cand[i,]
  cand[i,(cols+1)] <- names(which.max(abs(bar[4:cols]))) # gives the variable
  cand[i,(cols+2)] <- max(abs(bar[4:cols]))              # gives the correlation
}

colnames(cand)[cols+1] <- "predictor"
colnames(cand)[cols+2] <- "correlation"

table(cand$predictor)
# 266: bio 18 = Precipitation of Warmest Quarter
# 583 bio 2 = Mean Diurnal Temperature Range (Mean of monthly (max temp - min temp))
# 234 bio 7 = Temperature Annual Range
# 304 bio 8 = Mean Temperature of Wettest Quarter

write.table(cand,"ind.rda.6.memps.cand.snps.txt",row.names=F,sep = "\t", quote=F)

sel <- cand$snp
env <- cand$predictor
env[env=="bio18"] <- '#1b9e77'
env[env=="bio"] <- '#d95f02'
env[env=="bio7"] <- "purple4"
env[env=="bio8"] <-  '#e7298a'

col.pred <- rownames(losh.subset.ind.rda.memps$CCA$v) # pull the SNP names

for (i in 1:length(sel)) {           # color code candidate SNPs
  foo <- match(sel[i],col.pred)
  col.pred[foo] <- env[i]
}

col.pred[grep("scaffold",col.pred)] <- '#f1eef6' # non-candidate SNPs
col.pred[!grepl("^#", col.pred)] <- "#f1eef6"
empty <- col.pred
empty[grep("#f1eef6",empty)] <- rgb(0,1,0, alpha=0) # transparent

empty <- col.pred
empty[empty == "#f1eef6"] <- rgb(0, 1, 0, alpha = 0)

empty.outline <- ifelse(empty == "#00FF0000", "#00FF0000", "gray32")

bg <- c('#1b9e77','#d95f02',"purple4",'#e7298a')


plot(losh.subset.ind.rda.memps, type="n", scaling=3, xlim=c(-1,1), ylim=c(-1,1))
#points(cawa2.rda, display="species", pch=21, cex=1, col="gray32", bg=col.pred, scaling=3)
points(losh.subset.ind.rda.memps, display="species", pch=21, cex=1, col=empty.outline, bg=empty, scaling=3)
text(losh.subset.ind.rda.memps, scaling=3, display="bp", col="#0868ac", cex=1)
legend("bottomright", legend=c("bio18","bio2","bio07","bio8"), bty="n", col="gray32", pch=21, cex=1, pt.bg=bg)

```



